\chapter{Task lists for the user tests}\label{appendix:tasklists}

\section{Task list 1: testing insight and usability}\label{appendix:tasklists:prototype1}

The user is given some context, i.e., the user knows he/she is using a recommender system to find new music and he/she has a number artists in his/her artist library.

\begin{enumerate}
	\item Answer the following questions without interacting with the visualization:
		\begin{enumerate}
			\item Describe what you see. Which visual elements stand out? Which general structures can be identified?
			\item What do you think the visualization does?
			\item Which the elements of the user interface, do you think allow interaction?
			\item What do you think will happen when you:
				\begin{itemize}
					\item hover over an node of the graph?
					\item hover over one of the users?
					\item click on an item?
					\item click on a user?
				\end{itemize}
		\end{enumerate}
	\item Try to interact with the visualization. Answer the following questions:
		\begin{enumerate}
			\item Which of the artists displayed in the graph are artist suggestions?
			\item What are the links or edges in the visualization?
			\item	Suppose you want to add an item to your profile, what steps would you undertake?
		\end{enumerate}
	\item Add an item to your profile. Answer the following questions:
		\begin{enumerate}
			\item Why did you choose that particular item?
			\item Can you give any other reasons why you should pick this item?
			\item Can you give reasons for choosing one of the other items?
		\end{enumerate}
\end{enumerate}



\section{Task list 2: testing the first version of the settings menu}\label{appendix:tasklists:prototype2}

\begin{enumerate}
	\item Change the number of shown recommendations up to $20$.
	\item Change the threshold to $0.3$.
	\item Change the colours to a encoding that you like.
\end{enumerate}


\section{Task list 3: testing the performance of the evaluation system}\label{appendix:tasklists:prototype3}

\begin{enumerate}
	\item Find three neighbours that are closely related to you, based on the visualization.
	\item Find three recommended artists you think are interesting.
	\item Explain the recommendation rationale (transparency).
	\item Find a suggestion for an artist you didn't know about.
		\begin{itemize}
			\item Would you like to check our this artist's profile and listen one or more songs by this artist (persuasion)?
			\item Do you think the recommender system has made a good suggestion? Would you add it your profile (effectiveness)?
			\item How does it affect your trust in the recommender system (trust)?
		\end{itemize}
\end{enumerate}




